### 2018-08-30 abstract.tex
Teddy's comments implemented #20

Today climate change pushes humankind in search new ways to generate 
carbon-free, reliable base load power. Thus,
interest in advanced nuclear energy and particularly \glspl{MSR} has resurged, 
with multiple new companies pursuing commercialization of \gls{MSR} 
designs (e.g. Transatomic, Terrapower, Terrestrial, Moltex Energy, Thorcon). To further 
develop these \gls{MSR} concepts, researchers need simulation tools for analyzing liquid 
fueled molten salt reactor depletion and fuel processing. However, most 
contemporary nuclear reactor physics software is unable to perform high-fidelity 
full-core depletion calculations in an online reprocessing regime. This paper 
introduces a Python package, SaltProc, which couples with the Monte Carlo code, 
SERPENT2, and simulates \gls{MSR} online reprocessing by modeling the changing 
isotopic composition of \gls{MSR} fuel salt. SaltProc capabilities were 
demonstrated for full-core high-fidelity model of the commercial \gls{MSBR} 
concept and compared well to previous results from lower fidelity analyses.

>>>>

Today climate change pushes humankind in search of new ways to generate 
carbon-free, reliable base load power. Thus,
interest in advanced nuclear energy and particularly \glspl{MSR} has resurged, 
with multiple new companies pursuing commercialization of \gls{MSR} 
designs (e.g. Transatomic, Terrapower, Terrestrial, Moltex Energy, Thorcon). To further 
develop these \gls{MSR} concepts, researchers need simulation tools for analyzing liquid 
fueled molten salt reactor depletion and fuel processing. However, most 
contemporary nuclear reactor physics software is unable to perform high-fidelity 
full-core depletion calculations for a reactor design with online reprocessing. This paper 
introduces a Python package, SaltProc, which couples with the Monte Carlo code, 
SERPENT2, to simulate \gls{MSR} online reprocessing by modeling the changing 
isotopic composition of \gls{MSR} fuel salt. SaltProc capabilities were 
demonstrated for full-core high-fidelity model of the commercial \gls{MSBR} 
concept and compared well to previous results from lower-fidelity analyses.

### 2018-08-30 conclusion.tex
Thesis --> work
Teddy's comments implemented #20

\subsection{Equilibrium state search}
The results of this study indicate that the effective multiplication factor 
slowly decreases from 1.075 and reaches 1.02 at equilibrium after approximately 
6 years of operation. At the same time, the concentration of $^{233}$U, $^{232}$Th, 
$^{233}$Pa, $^{232}$Pa stabilizes after approximately 2500 days of operation. 
Particularly, $^{233}$U number density equilibrates\footnote{fluctuates less 
than 0.8\%} after 16 years of operation. Consequently, the core reaches the quasi-equilibrium state 
after 16 years of operation. On the other hand, a wide diversity of nuclides, 
including fissile isotopes (e.g. $^{233}$U, $^{239}$Pu) and non-fissile strong 
absorbers (e.g. $^{234}$U), continue accumulating in the core. The current work results 
show that a true equilibrium composition cannot exist but the balance
between strong absorber accumulation and new fissile material 
production can be achieved to keep the reactor critical.

@@

For thermal spectrum \gls{MSBR} removal of volatile gases, noble metals, and 
rare earths from a fuel salt benefits core performance.
Moreover, immediate removal of volatile gases (e.g., xenon) and noble metals 
increased reactivity by approximately 7500 pcm over a 10-year 
timeframe. In contrast, the effect of relatively slower removal of rare earth 
elements (every 50 days cycle instead of 3 days) has less impact (5500 pcm) on 
the core reactivity after 10 years of operation. In sum, additional study 
is needed to establish neutronic  and economic tradeoffs of removing each element.

@@

Only the batch-wise online reprocessing approach has been treated in this 
thesis. However, the SERPENT 2 Monte Carlo code was recently extended for 
continuous online fuel reprocessing simulation \cite{aufiero_extended_2013}. 
This extension must be verified against existing SaltProc/SERPENT or 
ChemTriton/SCALE packages, and could be employed for immediate removal of 
fission product gases (e.g., Xe, Kr) which have a strong negative impact on 
core lifetime and breeding efficiency. Finally, using the built-in SERPENT 2 
Monte Carlo code online reprocessing \& refueling material burnup routine would 
significantly speed up computer-intensive full-core depletion simulations.

>>>>

\subsection{Equilibrium state search}
The results of this study indicate that the effective multiplication factor 
slowly decreases from 1.075 and reaches 1.02 at equilibrium after approximately 
6 years of operation. At the same time, the concentration of $^{233}$U, $^{232}$Th, 
$^{233}$Pa, $^{232}$Pa stabilizes after approximately 2500 days of operation. 
Particularly, $^{233}$U number density equilibrates\footnote{fluctuates less 
than 0.8\%} after 16 years of operation. Consequently, the core reaches the quasi-equilibrium state 
after 16 years of operation. On the other hand, a wide variety of nuclides, 
including fissile isotopes (e.g. $^{233}$U, $^{239}$Pu) and non-fissile strong 
absorbers (e.g. $^{234}$U), continue accumulating in the core. The current work results 
show that a true equilibrium composition cannot exist but the balance
between strong absorber accumulation and new fissile material 
production can be achieved to keep the reactor critical.

@@

For thermal spectrum \gls{MSBR}, removal of volatile gases, noble metals, and 
rare earths from a fuel salt benefits core performance.
Moreover, immediate removal of volatile gases (e.g., xenon) and noble metals 
increased reactivity by approximately 7500 pcm over a 10-year 
timeframe. In contrast, the effect of relatively slower removal of rare earth 
elements (every 50 days cycle instead of 3 days) has less impact (5500 pcm) on 
the core reactivity after 10 years of operation. In sum, additional study 
is needed to establish neutronic  and economic tradeoffs of removing each element.

@@

Only the batch-wise online reprocessing approach has been treated in this 
work. However, the SERPENT 2 Monte Carlo code was recently extended for 
continuous online fuel reprocessing simulation \cite{aufiero_extended_2013}. 
This extension must be verified against existing SaltProc/SERPENT or 
ChemTriton/SCALE packages, and could be employed for immediate removal of 
fission product gases (e.g., Xe, Kr) which have a strong negative impact on 
core lifetime and breeding efficiency. Finally, using the built-in SERPENT 2 
Monte Carlo code online reprocessing \& refueling material burnup routine would 
significantly speed up computer-intensive full-core depletion simulations.

### 2018-08-30 methods.tex
Teddy's comments implemented #20

There are eight symmetric graphite slabs with a width of 15.24 cm in zone II, 
one of which is illustrated in Figure~\ref{fig:serpent_zoneII}. The holes in 
the centers are for the core lifting rods used during the core replacement 
operations. These holes also allow a portion of the fuel salt to flow to the 
top of the vessel for cooling the top head and axial reflector. 
Figure~\ref{fig:serpent_zoneII} also shows :w
the 5.08-cm-wide annular 
space between the removable core graphite in zone II-B and the permanently 
mounted reflector graphite. This annulus consists entirely of fuel salt, 
provides space for moving the core assembly, helps compensate for the elliptical 
dimensions of the reactor vessel, and serves to reduce the damaging flux at the 
surface of the graphite reflector blocks. In this work, all figures of the core 
were generated using the built-in SERPENT 2 plotter.

@@

The tool structure and capabilities of SaltProc are similar to the ChemTriton tool 
developed in \gls{ORNL} for SCALE \cite{powers_new_2013}. SaltProc is coupled 
with the Monte Carlo SERPENT 2
software to simulate online reprocessing for irregular full-core 
geometry with high fidelity.  The primary function of SaltProc is to 
manage material streams while
SERPENT 2 performs the computationally heavy work, namely neutron 
transport and depletion
calculations. Saltproc is defined as a python class, where each material stream 
is defined as a isotopic atomic density
vector variable. This allows tracking of time-sensitive material streams such 
as the
$^{233}$Pa tank in the \gls{MSBR}. The user can define the reprocessing 
parameters, such as the reprocessing interval and removal efficiency.  In 
addition, SaltProc provides a set of functions for each stream: read and write 
isotopic data in/from database, separate out specific isotopes from stream with 
defined efficiency, feed in specific isotopes to stream, and maintain constant 
number density of specific nuclide in the core. These attributes and functions 
are crucial to simulating the operation of a complex, multi-zone, multi-fluid 
\gls{MSR} and sufficiently general to represent myriad reactor systems.

SaltProc, currently in active development in Github (https://github.com/ 
arfc/saltproc), leverages unit tests and continuous integration for 
sustainable development. There is also documentation
generated through Sphinx document generator for ease of use. In future 
releases, we plan to 
implement
support for entirely user-customized reprocessing strategies, two-region \gls{MSR} modeling 
capabilities,
and decay modeling.

>>>>

There are eight symmetric graphite slabs with a width of 15.24 cm in zone II, 
one of which is illustrated in Figure~\ref{fig:serpent_zoneII}. The holes in 
the centers are for the core lifting rods used during the core replacement 
operations. These holes also allow a portion of the fuel salt to flow to the 
top of the vessel for cooling the top head and axial reflector. 
Figure~\ref{fig:serpent_zoneII} also shows
the 5.08-cm-wide annular 
space between the removable core graphite in zone II-B and the permanently 
mounted reflector graphite. This annulus consists entirely of fuel salt, 
provides space for moving the core assembly, helps compensate for the elliptical 
dimensions of the reactor vessel, and serves to reduce the damaging flux at the 
surface of the graphite reflector blocks. In this work, all figures of the core 
were generated using the built-in SERPENT 2 plotter.

@@

The tool structure and capabilities of SaltProc are similar to the ChemTriton tool 
developed in \gls{ORNL} for SCALE \cite{powers_new_2013}. SaltProc is coupled 
with the Monte Carlo SERPENT 2
software to simulate online reprocessing for irregular full-core 
geometry with high fidelity.  The primary function of SaltProc is to 
manage material streams while
SERPENT 2 performs the computationally heavy work, namely neutron 
transport and depletion
calculations. Saltproc is defined as a python class, where each material stream 
is defined as a isotopic atomic density
vector variable. This allows tracking of time-sensitive material streams such 
as the
$^{233}$Pa tank in the \gls{MSBR}. The user can define the reprocessing 
parameters, such as the reprocessing interval and removal efficiency.  In 
addition, SaltProc provides a set of functions for each stream: read and write 
isotopic data in/from database, separate out specific elements from stream with 
defined efficiency, feed in specific isotopes to stream, and maintain constant 
number density of specific nuclide in the core. These attributes and functions 
are crucial to simulating the operation of a complex, multi-zone, multi-fluid 
\gls{MSR} and are sufficiently general to represent myriad reactor systems.

SaltProc, currently in active development in Github (https://github.com/ 
arfc/saltproc), leverages unit tests and continuous integration for 
sustainable development. There is also documentation
generated through Sphinx document generator for ease of use. In future 
releases, we plan to 
implement
support for entirely user-customized reprocessing strategies, two-region \gls{MSR} modeling 
capabilities, and decay modeling in tanks.

### 2018-08-30 results.tex
Teddy's comments implemented #20

First, SERPENT calculates the effective multiplication factor for the beginning 
of cycle time (fresh fuel composition for the first step). Next, it computes 
the new fuel salt composition for the end of a 3-day depletion step. The 
corresponding effective multiplication factor is much smaller than the previous 
one. Finally, SERPENT calculates $k_{eff}$ for the depleted composition after 
applying feeds and removals. $K_{eff}$ this increases accordingly since major reactor 
poisons (e.g. Xe, Kr) are removed, while fresh fissile material ($^{233}$U) 
from the protactinium decay tank is added.

@@

Figure~\ref{fig:spectrum_zones} shows that zone I produced more thermal neutrons 
than zone II, corresponding to a majority of fissions occuring in the central part 
of the core. In the undermoderated zone II, the neutron energy spectrum is harder 
which leads to more capture of neutrons by $^{232}$Th and helps a achieve relatively 
high breeding ratio. Moreover, the (n,$\gamma$) resonance energy range in $^{232}$Th 
is from 10$^{-4}$ to 10$^{-2}$ MeV. Therefore, the moderator-to-fuel ratio for zone 
II was chosen to shift the neutron energy spectrum in this range. Furthermore, in the 
central core region (zone I), the neutron energy spectrum shifts to a harder spectrum 
over 20 years of reactor operation. Meanwhile, in the outer core region (zone II) a 
similar spectral shift takes place at a reduced scale. This results is in a good 
agreement with original ORNL report \cite{robertson_conceptual_1971} and most recent 
whole-core steady-state study \cite{park_whole_2015}.

@@

\subsection{Neutron flux}
Figure~\ref{fig:radial_flux} shows the radial distribution of fast and thermal 
neutron flux for both initial and equilibrium composition. The neutron flux has 
the same shape for both compositions but the equilibrium case has a harder 
spectrum. A significant spectral shift was observed for the central region of 
the core (zone I) when for the outer region (zone II) it is negligible for fast 
but notable for thermal neutrons. This neutron flux radial distribution is in a 
good agreement with original ORNL report \cite{robertson_conceptual_1971}. 
Overall, spectrum hardening during \gls{MSBR} operation should be carefully 
studied for designing the reactivity control system.
\begin{figure}[ht!] % replace 't' with 'b' to force it to \centering
  \includegraphics[width=\textwidth]{radial_flux.png} \caption{Radial neutron 
  flux distribution for initial and equilibrium fuel salt composition.}
  \label{fig:radial_flux}
\end{figure}

@@

initial and equilibrium fuel composition. Figure~\ref{fig:pow_den} reflects the 

@@

\subsection{Reactivity control system rod worth}
Table~\ref{tab:rod_worth} summarizes the reactivity control system worth. 
During normal operation the control (graphite) rods are fully inserted, and the 
safety (B$_4$C) rods are fully withdrawn. To insert negative reactivity into 
the core, the graphite rods are gradually withdrawn from the core. In an 
accident, the safety rods would fall down into the core. The integral rod 
worths were calculated for various positions to separately estimate control 
graphite moderator rods\footnote{In \cite{robertson_conceptual_1971}, the 
graphite rods are referred to as ``control'' rods.}, safety (B$_4$C) rod, and 
the whole reactivity control system worth. Control rod integral worth is 
approximately 28 cents and stays almost constant during reactor operation. The 
safety rod integral worth decreases by  16.2\% during 20 years of operation 
because of neutron spectrum hardening and absorber accumulation in proximity to 
reactivity control system rods. This 16\% decline in control system worth 
should be taken into account in \gls{MSBR} accident analysis and safety 
justification.
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\begin{table}[ht!]
  \centering
  \caption{Control system rod worth for initial and equilibrium fuel 
  composition.}
\begin{tabularx}{\textwidth}{ b | x | x } \hline
Reactivity parameter [cents]  &  Initial      &  Equilibrium      \\ \hline
Control (graphite) rod integral worth               & $\ 28.2\pm0.8$    & $\ 
        29.0\pm0.8$ \\ Safety (B$_4$C) rod integral worth                  & 
        $251.8\pm0.8$    & $211.0\pm0.8$  \\
Total reactivity control system worth               & $505.8\pm0.7$    & 
        $424.9\pm0.8$ \\ \hline
\end{tabularx}
  \label{tab:rod_worth}
\end{table}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

@@

\subsection{The effect of removing fission product from fuel salt}
Loading initial fuel salt composition into the \gls{MSBR} core leads to 
supercritical configuration (Figure ~\ref{fig:fp_removal}). After reactor 
startup the effective multiplication factor for the case with volatile gases 
and noble metals removal is approximately 7500 pcm  higher than for case with 
no fission products removal. This significant impact on the reactor core 
achieved due to immediate removal (20 sec cycle time) and high absorption cross 
section of Xe, Kr, Mo, and other noble metals removed. The effect of rare earth 
element removal considerable after few month from startup and achieves 
approximately 5500 pcm after 10 years of operation. The rare earth elements are 
removed with slower rate (50 days cycle time). Moreover, 
Figure~\ref{fig:fp_removal} demonstrates that batch-wise removal every 3-day 
step even strong absorbers did not necessarily leads to fluctuation in results 
but rare earth elements removal every 50 days causes approximately 600 pcm jump 
in reactivity.

The effective multiplication factor of the core reduces gradually over 
operation time because the fissile material ($^{233}$U) continuously depletes 
from the fuel salt due to fission and, at the same time, fission products 
accumulates in the fuel salt.	Eventually, without fission products removal, 
the reactivity decreases to the subcritical state after approximately 500 and 
1300 days of operation for cases with no removal and volatile gases \& noble 
metals removal, respectively. The time when the simulated core reaches 
subcriticality ($k_{eff}<$1.0) for full-core model) called the core lifetime. 
Therefore, removing fission products provides with significant neutronic benefit 
and enables a longer core lifetime.
\begin{figure}[ht!] % replace 't' with 'b' to force it to 
  \centering
  \includegraphics[width=\textwidth]{keff_rem_cases.png} 
  \caption{Calculated effective multiplication factor for full-core \gls{MSBR} 
model with removal of various fission product groups over 10 years of 
operation.}
  \label{fig:fp_removal}
\end{figure}

>>>>

First, SERPENT calculates the effective multiplication factor for the beginning 
of cycle time (fresh fuel composition for the first step). Next, it computes 
the new fuel salt composition at the end of a 3-day depletion step. The 
corresponding effective multiplication factor is much smaller than the previous 
one. Finally, SERPENT calculates $k_{eff}$ for the depleted composition after 
applying feeds and removals. This $K_{eff}$ increases accordingly since major reactor 
poisons (e.g. Xe, Kr) are removed, while fresh fissile material ($^{233}$U) 
from the protactinium decay tank is added.

@@

Figure~\ref{fig:spectrum_zones} shows that zone I produced more thermal neutrons 
than zone II, corresponding to a majority of fissions occuring in the central part 
of the core. In the undermoderated zone II, the neutron energy spectrum is harder 
which leads to more capture of neutrons by $^{232}$Th and helps a achieve relatively 
high breeding ratio. Moreover, the (n,$\gamma$) resonance energy range in $^{232}$Th 
is from 10$^{-4}$ to 10$^{-2}$ MeV. Therefore, the moderator-to-fuel ratio for zone 
II was chosen to shift the neutron energy spectrum in this range. Furthermore, in the 
central core region (zone I), the neutron energy spectrum shifts to a harder spectrum 
over 20 years of reactor operation. Meanwhile, in the outer core region (zone II) a 
similar spectral shift takes place at a reduced scale. These results are in a good 
agreement with original ORNL report \cite{robertson_conceptual_1971} and the most recent 
whole-core steady-state study \cite{park_whole_2015}.

@@

\subsection{Neutron flux}
Figure~\ref{fig:radial_flux} shows the radial distribution of fast and thermal 
neutron flux for the both initial and equilibrium composition. The neutron flux has 
the same shape for both compositions but the equilibrium case has a harder 
spectrum. A significant spectral shift was observed for the central region of 
the core (zone I) when for the outer region (zone II) it is negligible for fast 
but notable for thermal neutrons. This neutron flux radial distribution is in a 
good agreement with original ORNL report \cite{robertson_conceptual_1971}. 
Overall, spectrum hardening during \gls{MSBR} operation should be carefully 
studied for designing the reactivity control system.
\begin{figure}[ht!] % replace 't' with 'b' to force it to \centering
  \includegraphics[width=\textwidth]{radial_flux.png} \caption{Radial neutron 
  flux distribution for initial and equilibrium fuel salt composition.}
  \label{fig:radial_flux}
\end{figure}

@@

initial and equilibrium fuel compositions. Figure~\ref{fig:pow_den} reflects the 

@@

\subsection{Reactivity control system rod worth}
Table~\ref{tab:rod_worth} summarizes the reactivity control system worth. 
During normal operation the control (graphite) rods are fully inserted, and the 
safety (B$_4$C) rods are fully withdrawn. To insert negative reactivity into 
the core, the graphite rods are gradually withdrawn from the core. In an 
accident, the safety rods would be dropped down into the core. The integral rod 
worths were calculated for various positions to separately estimate control 
graphite moderator rods\footnote{In \cite{robertson_conceptual_1971}, the 
graphite rods are referred to as ``control'' rods.}, safety (B$_4$C) rod, and 
the whole reactivity control system worth. Control rod integral worth is 
approximately 28 cents and stays almost constant during reactor operation. The 
safety rod integral worth decreases by  16.2\% during 20 years of operation 
because of neutron spectrum hardening and absorber accumulation in proximity to 
reactivity control system rods. This 16\% decline in control system worth 
should be taken into account in \gls{MSBR} accident analysis and safety 
justification.
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\begin{table}[ht!]
  \centering
  \caption{Control system rod worth for initial and equilibrium fuel 
  composition.}
\begin{tabularx}{\textwidth}{ b | x | x } \hline
Reactivity parameter [cents]  &  Initial      &  Equilibrium      \\ \hline
Control (graphite) rod integral worth               & $\ 28.2\pm0.8$    & $\ 
        29.0\pm0.8$ \\ Safety (B$_4$C) rod integral worth                  & 
        $251.8\pm0.8$    & $211.0\pm0.8$  \\
Total reactivity control system worth               & $505.8\pm0.7$    & 
        $424.9\pm0.8$ \\ \hline
\end{tabularx}
  \label{tab:rod_worth}
\end{table}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

@@

\subsection{The effect of removing fission product from fuel salt}
Loading initial fuel salt composition into the \gls{MSBR} core leads to 
supercritical configuration (Figure ~\ref{fig:fp_removal}). After reactor 
startup, the effective multiplication factor for the case with volatile gases 
and noble metals removal is approximately 7500 pcm  higher than for case with 
no fission products removal. This significant impact on the reactor core is
achieved due to immediate removal (20 sec cycle time) and high absorption cross 
section of Xe, Kr, Mo, and other noble metals removed. The effect of rare earth 
element removal is considerable after few month from startup and achieves 
approximately 5500 pcm after 10 years of operation. The rare earth elements are 
removed with a slower rate (50 days cycle time). Moreover, 
Figure~\ref{fig:fp_removal} demonstrates that batch-wise removal every 3-day 
step even strong absorbers did not necessarily leads to fluctuation in results 
but rare earth elements removal every 50 days causes an approximately 600 pcm jump 
in reactivity.

The effective multiplication factor of the core reduces gradually over 
operation time because the fissile material ($^{233}$U) continuously depletes 
from the fuel salt due to fission and, at the same time, fission products 
accumulate in the fuel salt.	Eventually, without fission products removal, 
the reactivity decreases to the subcritical state after approximately 500 and 
1300 days of operation for cases with no removal and volatile gases \& noble 
metals removal, respectively. The time when the simulated core reaches 
subcriticality ($k_{eff}<$1.0) for full-core model) is called the core lifetime. 
Therefore, removing fission products provides with significant neutronic benefit 
and enables a longer core lifetime.
\begin{figure}[ht!] % replace 't' with 'b' to force it to 
  \centering
  \includegraphics[width=\textwidth]{keff_rem_cases.png} 
  \caption{Calculated effective multiplication factor for full-core \gls{MSBR} 
model with removal of various fission product groups over 10 years of 
operation.}
  \label{fig:fp_removal}
\end{figure}
